PhD Thesis

### 2016-12-02 aosd13/Evaluation.tex
aosd13: adjust relative paths
Replace paper with chapter (with care)

\newcommand{\graphPath}[1]{graphs/#1}
\input{graphs/evalRes}
\section{Evaluation}
\label{sec:evaluation}

@@

\subsection{Study Setup}
%evaluation strategy
Throughout the paper, we have already shown several compact queries for which our optimizations increase performance significantly compared to a naive execution. Since some optimizations change 
the complexity class of the query (e.g.\ by using an index), so the speedups grow with the size of the data. However, to get a more realistic evaluation of \LoS{}, we decided to perform an experiment with existing real-world queries.

@@

\begin{figure}
	\centering
		\includegraphics[width=\linewidth]{graphs/measurements-overview}
	\caption{Measurement Setup: Overview}
	\label{fig:measurements-overview}
\end{figure}

@@

\input{graphs/queryTables}

@@

\begin{center}
\begin{tabular}{l*{1}{r@{}c@{}l}}\toprule
Index&\multicolumn{3}{c}{Elapsed time (ms)}\\\midrule
\input{graphs/indexes}
\bottomrule
\end{tabular}
\end{center}
For our test data, index construction takes less than 200 ms for the first two indexes, which is moderate compared to the time for loading the bytecode in the BAT representation ($\readingClassFilesTime$). Building the instruction index took around 4 seconds, which we consider acceptable since this index maps each type of instruction (e.g.\ \code{INSTANCEOF}) to a collection of all bytecode instructions of that type.
%Valid question: why does the exception-handler index take so much more than the query it optimizes? That's because indexes are interpreted. But it's best not to say it. Also, we already discuss this later convincingly.

>>>>

\newcommand{\graphPath}[1]{aosd13/graphs/#1}
\input{aosd13/graphs/evalRes}
\section{Evaluation}
\label{sec:evaluation}

@@

\subsection{Study Setup}
%evaluation strategy
Throughout this chapter, we have already shown several compact queries for which our optimizations increase performance significantly compared to a naive execution. Since some optimizations change
the complexity class of the query (e.g.\ by using an index), so the speedups grow with the size of the data. However, to get a more realistic evaluation of \LoS{}, we decided to perform an experiment with existing real-world queries.

@@

\begin{figure}
	\centering
		\includegraphics[width=\linewidth]{aosd13/graphs/measurements-overview}
	\caption{Measurement Setup: Overview}
	\label{fig:measurements-overview}
\end{figure}

@@

\input{aosd13/graphs/queryTables}

@@

\begin{center}
\begin{tabular}{l*{1}{r@{}c@{}l}}\toprule
Index&\multicolumn{3}{c}{Elapsed time (ms)}\\\midrule
\input{aosd13/graphs/indexes}
\bottomrule
\end{tabular}
\end{center}
For our test data, index construction takes less than 200 ms for the first two indexes, which is moderate compared to the time for loading the bytecode in the BAT representation ($\readingClassFilesTime$). Building the instruction index took around 4 seconds, which we consider acceptable since this index maps each type of instruction (e.g.\ \code{INSTANCEOF}) to a collection of all bytecode instructions of that type.
%Valid question: why does the exception-handler index take so much more than the query it optimizes? That's because indexes are interpreted. But it's best not to say it. Also, we already discuss this later convincingly.

### 2016-12-02 aosd13/paperBody.tex
aosd13: adjust relative paths
Drop duplicated label
Replace paper with chapter (with care)

As our running example through the paper, we consider representing and querying a simple in-memory bibliography. A book has, in our schema, a title, a publisher and a list of authors. Each author, in turn, has a first and last name. We represent authors and books as instances of the Scala classes \code{Author} and \code{Book} shown in Fig.~\ref{fig:schema}. The class declarations list the type of each field: Titles, publishers, and first and last names are all stored in fields of type \code{String}. The list of authors is stored in a field of type \code{Seq[Author]}, that is, a sequence of authors -- something that would be more complex to model in a relational database.
The code fragment also defines a collection of books named \code{books}.
\begin{figure}[t]
\centering
\begin{lstlisting}
package schema
case class Author(firstName: String, lastName: String)
case class Book(title: String, publisher: String,
  authors: Seq[Author])

@@

\input{Evaluation}

\section{Related Work}
\label{sec:relwork}
This paper builds on prior work on language-integrated queries, query optimization, techniques for DSL embedding, and other works on code querying.

@@

There have been many approaches for a closer integration of SQL queries into programs, such as
HaskellDB~\citep{Leijen99DSEC} (which also inspired \LINQ), or Ferry~\citep{Grust:2009:FDP:1559845.1559982} 
(which moves part of a program execution to a database). In Scala, there are also
APIs which integrate SQL queries more closely such as
Slick.\footnote{\url{http://slick.typesafe.com/}} Its frontend allows to define
and combine type-safe queries, similarly to ours (also in the way it is
implemented).
However, the language for defining queries maps to SQL, so it does not support nesting collections
in other collections (a feature which simplified our example in
Sec.~\ref{sec:motivation}), nor distinguishes statically between different kinds of
collections, such as \code{Set} or \code{Seq}.
Based on Ferry, ScalaQL~\citep{JOT:issue_2010_07/article3} extends Scala with a compiler-plugin to integrate a query language on top of a relational database. The work by \citet{Spiewak09scalaql:language-integrated} is
 unrelated to~\citep{JOT:issue_2010_07/article3} but also called ScalaQL\@. It is similar to our approach in that it also 
proposes to reify queries based on for-comprehensions, but it is not clear from the paper how the reification 
works.\footnote{We contacted the authors; they were not willing to provide more details or the sources of their approach.}

@@

\smartParagraph{Scala and DSL Embedding}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling}. The technical
details of the embedding are not the main topic of this paper; we are using some of the 
Scala techniques presented by \citet{rompf2010lightweight} for using implicits and for
adding infix operators to a type. Similar to \citet{rompf2010lightweight}, we also 
use the Scala compiler on-the-fly. A plausible alternative backend for \LoS\ would
have been to use Delite~\citep{Rompf11BBlocks}, a framework for
building highly efficient DSLs in Scala.
Using this framework, in concurrent work, \citet{Rompf13} also optimize
collection queries; while their work allows for imperative programs, they do
not support embedding arbitrary libraries in an automated way. On the other
hand, they can reuse support for automatic parallelization and multiple platforms present in Delite.
\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Moreover, both works do not apply typical database optimizations such as
indexing or filter hoisting.

@@

\section{Conclusions}
\label{sec:concl}

>>>>

As our running example through the chapter, we consider representing and querying a simple in-memory bibliography. A book has, in our schema, a title, a publisher and a list of authors. Each author, in turn, has a first and last name. We represent authors and books as instances of the Scala classes \code{Author} and \code{Book} shown in Fig.~\ref{fig:schema}. The class declarations list the type of each field: Titles, publishers, and first and last names are all stored in fields of type \code{String}. The list of authors is stored in a field of type \code{Seq[Author]}, that is, a sequence of authors -- something that would be more complex to model in a relational database.
The code fragment also defines a collection of books named \code{books}.
\begin{figure}[t]
\centering
\begin{lstlisting}
package schema
case class Author(firstName: String, lastName: String)
case class Book(title: String, publisher: String,
  authors: Seq[Author])

@@

\input{aosd13/Evaluation}

\section{Related Work}
\label{sec:relwork}
This chapter builds on prior work on language-integrated queries, query optimization, techniques for DSL embedding, and other works on code querying.

@@

There have been many approaches for a closer integration of SQL queries into programs, such as
HaskellDB~\citep{Leijen99DSEC} (which also inspired \LINQ), or Ferry~\citep{Grust:2009:FDP:1559845.1559982} 
(which moves part of a program execution to a database). In Scala, there are also
APIs which integrate SQL queries more closely such as
Slick.\footnote{\url{http://slick.typesafe.com/}} Its frontend allows to define
and combine type-safe queries, similarly to ours (also in the way it is
implemented).
However, the language for defining queries maps to SQL, so it does not support nesting collections
in other collections (a feature which simplified our example in
Sec.~\ref{sec:motivation}), nor distinguishes statically between different kinds of
collections, such as \code{Set} or \code{Seq}.
Based on Ferry, ScalaQL~\citep{JOT:issue_2010_07/article3} extends Scala with a compiler-plugin to integrate a query language on top of a relational database. The work by \citet{Spiewak09scalaql:language-integrated} is
 unrelated to~\citep{JOT:issue_2010_07/article3} but also called ScalaQL\@. It is similar to our approach in that it also 
proposes to reify queries based on for-comprehensions, but it is not clear from their paper how the reification
works.\footnote{We contacted the authors; they were not willing to provide more details or the sources of their approach.}

@@

\smartParagraph{Scala and DSL Embedding}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling}. \pg{This makes no sense for the thesis.}
The technical
details of the embedding are not the main topic of this chapter; we are using some of the
Scala techniques presented by \citet{rompf2010lightweight} for using implicits and for
adding infix operators to a type. Similar to \citet{rompf2010lightweight}, we also 
use the Scala compiler on-the-fly. A plausible alternative backend for \LoS\ would
have been to use Delite~\citep{Rompf11BBlocks}, a framework for
building highly efficient DSLs in Scala.
Using this framework, in concurrent work, \citet{Rompf13} also optimize
collection queries; while their work allows for imperative programs, they do
not support embedding arbitrary libraries in an automated way. On the other
hand, they can reuse support for automatic parallelization and multiple platforms present in Delite.
\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Moreover, both works do not apply typical database optimizations such as
indexing or filter hoisting.

@@

\section{Conclusions}

### 2016-12-02 aosd13/pepm13/caseStudyDiscussion.tex
Replace paper with chapter (with care)

We also consider significant that Scala 2.8 required making both type inference and implicit resolution more powerful, specifically in order to support the collection library~\citep[Sec 21.7]{Moors10TCP,Odersky11book}; further extensions would be possible and desirable.
For instance, if type inference were extended with higher-order unification\footnote{\url{https://issues.scala-lang.org/browse/SI-2712}}~\citep{Pfenning88}, it would better support a part of our DSL interface (not discussed in this paper) by removing the need for type annotations.
%\pg{That's type indexing, to describe earlier}.

>>>>

We also consider significant that Scala 2.8 required making both type inference and implicit resolution more powerful, specifically in order to support the collection library~\citep[Sec 21.7]{Moors10TCP,Odersky11book}; further extensions would be possible and desirable.
For instance, if type inference were extended with higher-order unification\footnote{\url{https://issues.scala-lang.org/browse/SI-2712}}~\citep{Pfenning88}, it would better support a part of our DSL interface (not discussed in this chapter) by removing the need for type annotations.
%\pg{That's type indexing, to describe earlier}.

### 2016-12-02 pldi14/sec-change-structures.tex
Replace paper with chapter (with care)

This definition is interesting because we \emph{can} define the
change structure, but we have issues with its elimination forms.
Take a pair of functions $p = (f, g)$ and its change
$dp = \mathit{inj}_3 db$ (with $db : \Delta b)$. How are we to
compute the derivative of its first projection, $fst' p dp$? This
must be the nil change of $f$, but this has a few issues:
\begin{itemize}
\item such a derivative is not self-maintainable because it
  requires knowing $f$, so it is inefficient in the framework
  present in this paper; however, we are addressing this issue.
\item The result is an inefficient nil change, because we do not
  use a computed derivative but simply
  $\NilC{f} = \Diff{f}{f} = \Lam{x}{\Lam{dx}{\Diff{\App{f}{\Update
          x {\D{x}}}}{\App{f}{x}}}}$. One can imagine a few ways
  to avoid that, but some require nontrivial changes to the
  framework. For instance, one could add derivatives as an
  additional observation of basic functions, so that whenever we
  have access to a function we also have access to its
  derivative. We leave investigation on this point for future
  work.
\item The semantics of a pair or of its two elements together
  should arguably be equivalent by general reasons --- simply
  because the environment is itself a sort of product (that's for
  instance guaranteed by categorical semantics). This suggests
  that, if we allow omitting changes for components of products,
  we should also allow omitting them for entries in the context.
\item It is currently sometimes hard to detect whether a change
  is nil or not. We should arguably add a suitable method to
  change structures (which is specified to give a conservative
  answer), and implement it for function changes through yet
  another observation.
\end{itemize}

>>>>

This definition is interesting because we \emph{can} define the
change structure, but we have issues with its elimination forms.
Take a pair of functions $p = (f, g)$ and its change
$dp = \mathit{inj}_3 db$ (with $db : \Delta b)$. How are we to
compute the derivative of its first projection, $fst' p dp$? This
must be the nil change of $f$, but this has a few issues:
\begin{itemize}
\item such a derivative is not self-maintainable because it
  requires knowing $f$, so it is inefficient in the framework
  present in this chapter; however, we are addressing this issue.\pg{Update}
\item The result is an inefficient nil change, because we do not
  use a computed derivative but simply
  $\NilC{f} = \Diff{f}{f} = \Lam{x}{\Lam{dx}{\Diff{\App{f}{\Update
          x {\D{x}}}}{\App{f}{x}}}}$. One can imagine a few ways
  to avoid that, but some require nontrivial changes to the
  framework. For instance, one could add derivatives as an
  additional observation of basic functions, so that whenever we
  have access to a function we also have access to its
  derivative. We leave investigation on this point for future
  work.
\item The semantics of a pair or of its two elements together
  should arguably be equivalent by general reasons --- simply
  because the environment is itself a sort of product (that's for
  instance guaranteed by categorical semantics). This suggests
  that, if we allow omitting changes for components of products,
  we should also allow omitting them for entries in the context.
\item It is currently sometimes hard to detect whether a change
  is nil or not. We should arguably add a suitable method to
  change structures (which is specified to give a conservative
  answer), and implement it for function changes through yet
  another observation.
\end{itemize}

### 2016-12-02 pldi14/sec-intro.tex
Replace paper with chapter (with care)

To support automatic incrementalization, in this paper we introduce the \ILC\
(incrementalizing $\Gl$-calculi) framework. We define
an automatic program transformation $\DERIVE$
that \emph{differentiates} programs, that is, computes their
derivatives; $\DERIVE$ guarantees that
\begin{equation}
  \label{eq:correctness}
\App{f}{\Apply*{\D a}{a}}
\cong
\Apply{\App*{\App{\Derive{f}}{a}}{\D a}}{\App*{f}{a}}.
\end{equation}
where
$\cong$ is denotational equality,
$\D a$ is a change on $a$ and $\Apply{\D a}{a}$ denotes $a$
updated with change $\D a$, that is, the updated input of $f$.
Hence, we can optimize programs by replacing the left-hand side,
which recomputes the output from scratch, with the right-hand
side, which computes the output incrementally using derivatives.
% KO: I think this forward references confuses more than it helps.
%Our approach relates to \emph{finite differencing} but has a more
%general theory and support for first-class functions (see
%\cref{sec:finite-diff}).

@@

This paper makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which is more
  general than other work in the field because changes are
  first-class entities, they are distinct from base values and
  they are defined also for functions (\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for
pure $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that requires no run-time
support. The transformation produces an incremental program in the same language;
all optimization techniques for the original program are
applicable to the incremental program as well.
%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
%Since our incremental programs use no impure features, they are
%especially amenable to further optimizations, making this approach
%very suitable for further research.
%
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
We prove that our incrementalizing transformation $\DERIVE$
is correct~(\cref{eq:correctness})
by a machine-checked formalization in Agda~\citep{agda-head}.
The proof gives insight into the definition of $\DERIVE$: we
first construct the derivative $\EvalInc{-}$ of the denotational
semantics of a simply-typed $\lambda$-calculus term, that is, its
\emph{change semantics}.
%
Then, we show that $\DERIVE$ is produced by erasing
$\EvalInc{-}$ to a simply-typed program (\cref{sec:correctness}).

@@

%We mechanized the formalization, including the separation between
%core and plugins, in the dependently typed programming language
%Agda~\cite{agda-head}.
Our Agda formalization, Scala implementation and benchmark
results are available at the URL
\url{http://inc-lc.github.io/}.
All lemmas and theorems presented
in this paper have been proven in Agda.
In the paper, we present an overview of
the formalization in more human-readable form, glossing over some
technical details.

>>>>

To support automatic incrementalization, in this chapter we introduce the \ILC\
(incrementalizing $\Gl$-calculi) framework. We define
an automatic program transformation $\DERIVE$
that \emph{differentiates} programs, that is, computes their
derivatives; $\DERIVE$ guarantees that
\begin{equation}
  \label{eq:correctness}
\App{f}{\Apply*{\D a}{a}}
\cong
\Apply{\App*{\App{\Derive{f}}{a}}{\D a}}{\App*{f}{a}}.
\end{equation}
where
$\cong$ is denotational equality,
$\D a$ is a change on $a$ and $\Apply{\D a}{a}$ denotes $a$
updated with change $\D a$, that is, the updated input of $f$.
Hence, we can optimize programs by replacing the left-hand side,
which recomputes the output from scratch, with the right-hand
side, which computes the output incrementally using derivatives.
% KO: I think this forward references confuses more than it helps.
%Our approach relates to \emph{finite differencing} but has a more
%general theory and support for first-class functions (see
%\cref{sec:finite-diff}).

@@

This chapter makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which is more
  general than other work in the field because changes are
  first-class entities, they are distinct from base values and
  they are defined also for functions (\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for
pure $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that requires no run-time
support. The transformation produces an incremental program in the same language;
all optimization techniques for the original program are
applicable to the incremental program as well.
%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
%Since our incremental programs use no impure features, they are
%especially amenable to further optimizations, making this approach
%very suitable for further research.
%
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
We prove that our incrementalizing transformation $\DERIVE$
is correct~(\cref{eq:correctness})
by a machine-checked formalization in Agda~\citep{agda-head}.
The proof gives insight into the definition of $\DERIVE$: we
first construct the derivative $\EvalInc{-}$ of the denotational
semantics of a simply-typed $\lambda$-calculus term, that is, its
\emph{change semantics}.
%
Then, we show that $\DERIVE$ is produced by erasing
$\EvalInc{-}$ to a simply-typed program (\cref{sec:correctness}).

@@

%We mechanized the formalization, including the separation between
%core and plugins, in the dependently typed programming language
%Agda~\cite{agda-head}.
Our Agda formalization, Scala implementation and benchmark
results are available at the URL
\url{http://inc-lc.github.io/}.
All lemmas and theorems presented
in this chapter have been proven in Agda.
In the chapter, we present an overview of
the formalization in more human-readable form, glossing over some
technical details.
